The Ex-Ante View of Recommender System Design

### 2019-08-28 paper/ex_ante_rec_sys.tex
add draft of paper to the repo

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage{mathtools}
\usepackage[font=small]{caption}
\usepackage{subcaption}
\usepackage{float}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

%
% end of the preamble, start of the body of the document source.
\begin{document}

@@

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
Recommender Systems (RS) are traditionally deployed in environments where users are uncertain about their preferences and thus face a problem of choice under uncertainty, but most popular design approaches ignore this fact. We argue that predicting and modeling consumer choice in these contexts can improve the usefulness of RS and reframe the problem as providing useful information to help reduce user uncertainty as opposed to simply predicting user preferences. We show how this insight can be utilized to design RS that mitigate the negative consequences such as filter bubble and user homogenization effects, as well as better understand the role that RS play in contributing to these phenomena.
\end{abstract}

@@

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

@@

Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube. On these platforms the combination of most goods being experience goods and the number of choices available to users being incredibly large, results in users not knowing their true preference ordering over all of the items. For instance, on Netflix there are thousands of movies, Amazon millions of products, and YouTube billions of videos. As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty. 
\par
However, while acknowledging that this is why RS are useful, most of the literature on RS ignores this user uncertainty and takes what we call an \textit{ex-post} viewpoint to preferences, focusing on the reported utility of users after they consume the item or inferring them via behavioral data \cite{adomavicius2005toward, zhao2018interpreting}. The RS problem traditionally is cast as solving the following prediction problem:
\begin{align*}
\arg\max\limits_{x \in A} u(x)
\end{align*}
\noindent where $A$ is the set of possible items and $u(x)$ represents the utility a user gets to consume good $x$. Recently there has been a move away from metrics focused on maximizing the accuracy of this prediction problem and towards optimizing for metrics such as serendipity or discoverability \cite{mcnee2006being, vargas2011rank} due to  the realization that accurate recommendations may not be the most useful recommendations for users. Furthermore, deployed RS have come under scrutiny for leading to ``filter bubbles" \cite{pariser2011filter} where users consume a less diverse set of content over time as well as increased homogenization amongst users \cite{chaney2018algorithmic, hosanagar2013will}.
\par
In this paper we want to put forward the argument that one approach to overcoming these problems is to make understanding how users choose which items to consume a first-order consideration when designing RS. In particular, we differentiate between the previous \textit{ex-post} interpretation of the RS problem and what we define as the \textit{ex-ante} viewpoint. Unlike the ex-post viewpoint taken in the traditional prediction problem, the ex-ante viewpoint simply states that users face uncertainty and so maximize their expected utility as is defined in economic decision theory under uncertainty \citep[see e.g.][]{mas1995microeconomic}:
\begin{align*}
\arg\max\limits_{x \in A} \mathbb{E}[u(x)]
\end{align*}
Fundamentally, the move away from accuracy towards alternative metrics as well as tensions that have arisen from increased user homogenization and filter bubbles come from the \textit{choices} that users make. This shift from thinking about true preferences after all uncertainty is resolved (the ex-post view) to user decisions under uncertainty (the ex-ante view) can serve as a theoretical tool to better understand what recommendations are useful for users as well as understanding what of the documented perverse effects of RS come from recommendation per-se and what effects come from inherent uncertainty in user decision-making.
\section{The Value of the Ex-Ante View}

In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS. Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make sub-optimal consumption decisions. Consider the following stylized example. The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
However when users are making decisions, they have unobservable \textit{beliefs} about the ex-post utility value of the items, and make decisions given these beliefs. If users had more information then their beliefs would converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values. In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 1)$.
In this example, after uncertainty is unresolved, the user ought to consume $x_3$, but an expected utility maximizing consumer (without further information) would consume $x_1$.\footnote{Note that here we ignore an important component of decision-making under uncertainty from decision theory, risk preferences, and suppose users are risk-neutral. A \textit{risk-averse} user may choose to consume a good with lower expected payoff but lower ex-ante variance compared to another good. While risk-aversion is an important component of decision-making and important in the context of RS when considering choices under uncertainty, it is not our main focus here.}
This observation, while seemingly trivial, has important implications. The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g. \cite[]{mas1995microeconomic}).} and allows us to get some information about a user's beliefs. Why are user beliefs important objects for designers of RS to understand and utilize?
The first is that they are useful for understanding how users interact and get value from RS. In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items. For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$. It is an empirical question that we leave for future work, precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}. Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}. Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations. For instance, in the example, the user does not consume $x_3$ because she believes that the good has the potential to be catastrophically bad but, given more information, she would prefer to consume it over $x_1$ and $x_2$. It is possible that the ex-post and ex-ante utility of $x_3$ could be at best as good as the other two. The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item but high expected quality.}
As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it. This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.\footnote{\cite{schotter2014belief} provides a survey of belief elicitation methods that are utilized in economics.}
Understanding the choices that individuals will make as well as what they will end up liking are complementary problems but require different viewpoints. \cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations. We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some of the results from \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user. For instance, on Netflix I may have seen every episode of Friends when it was on TV which influences the consumption choices that I make. Alternatively, my friends can be a source of information that affect beliefs and choices. Netflix cannot observe neither of these data.} 
Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected. In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem.

@@

Viewing the goal of the RS as providing users with additional information they can utilize to update their beliefs, we first show that, without recommendation, users may get stuck in a ``filter bubble" whereby they consume increasingly similar content and that, in this case, welfare will be negatively correlated with diversity. We show how providing information to users via recommendation can lead users out of this filter bubble. However, if recommendation does not properly take into account user beliefs, recommendations can lead to increasing homogenization among users and lower welfare and consumption diversity. Finally we then show that when the recommender does not know the user's beliefs about her valuation, recommendation becomes decreasingly useful as the user learns more about her own preferences.
\noindent \textbf{Setup}. Suppose that there is a group $I$ of individuals and that each individual $i \in I$ can choose from the same finite set of $N$ items $\mathcal{I} = \left\{0,1,...,N-1\right\}$. For simplicity, we assume that individuals only derive pleasure from a item $n \in \mathcal{I}$ the first time they consume it. We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$. We define the distance between two items as $d(n,m)=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{I}$ so that, conceptually, the items are located in a ring of size $N$.
In particular, we consider that the utility derived from a given item can be decomposed in the following manner: $u_{in}=\beta v_{in} + (1-\beta) v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e. consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. 
One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. 
The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=1$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta=0$, every individual has the same preferences.

@@

We assume that user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$.
Furthermore, we assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, i.e. $n\pm1$ (mod $N$), than about those farther away. This captures the idea that trying a product provides more information about similar products than about dissimilar ones.
To this effect, we consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.

@@

@@

We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare. The first will be measured by $D_i =\frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}. The second will be measured by a consumer homogeneity index $H:=1-\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard distance and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}. Note that a higher $H$ indicates more homogeneity. Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.

@@

Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$. In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility given beliefs at time $t$. Upon consuming that item and learning its value, the users update their beliefs about their valuation of other items and iterate this process $T$ times. In the second case, the recommender knows exactly $V_i$ and $V$ and can therefore recommend the best item for each consumer. In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$. The recommendation will be that the consumer chooses $r^{t} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$. For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{

@@

@@

\begin{table}[bt]
\centering
\begin{tabular}{rlll}
  \hline
Rec Policy & Welfare & Diversity & Homogeneity \\ 
  \hline
No Rec & 0.42 $\pm$ 0.002 & 191.04 $\pm$ 0.27 & 0.001 $\pm$ 0 \\ 
Partial Rec & 1.25 $\pm$ 0.005 & 221.83 $\pm$ 0.20 & 0.174 $\pm$ 0.004 \\ 
Omniscient Rec & 2.24 $\pm$ 0.006 & 237.59 $\pm$ 0.12 & 0.035 $\pm$ 0.002 \\ 
   \hline
\end{tabular}
\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
\label{table:agg_results}
\end{table}

\begin{figure*}[!htbp]
    \centering
    \begin{minipage}[b]{0.5\textwidth}
        \centering
        \includegraphics[width=\linewidth, height=1.75in]{figures/local_moves_25}
        \caption{Fraction of Local Search}
        \label{fig:local_consumption}
    \end{minipage}%
    ~ 
    \begin{minipage}[b]{0.5\textwidth}
        \centering
        \includegraphics[width=\linewidth, height=1.75in]{figures/welfare_homo_combo}
        \caption{Welfare and Homogeneity varying $\beta$}
        \label{fig:beta_vary}
    \end{minipage}
    ~ \\
     \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=\linewidth, height=1.75in]{figures/rec_obedience_25}
     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
   \end{minipage}\hfill
   ~
   \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=\linewidth, height=1.75in]{"figures/Diversity vs Welfare - No Recommendation"}
     \caption{ Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
   \end{minipage}
\end{figure*}
\small
\noindent \textbf{Results}. Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity. The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility. Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity. In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}. Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.  The partial recommendation case leads to higher consumption diversity and higher welfare, but has a significantly higher level of homogeneity.

In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products. It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this. The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
\par
Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity. As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation. If recommendation only provides information on the common-value component then welfare and consumption diversity increase but, correspondingly, homogenization increases dramatically. In this case the information on the common-value component leads users to have more information on where to explore in the product space. However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users. If recommendation takes into account user beliefs on the idiosyncratic  component of utility then welfare and consumption diversity can further increase and homogeneity will decrease.
\par
Low consumption diversity by itself is not necessarily bad. Figure \ref{fig:diversity_welfare_no_rec} shows that in the no recommendation case welfare and consumption diversity are negatively correlated. This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products. However, if a user finds a high utility product right away then staying in this neighborhood may yield higher utility due to the correlation of utilities. The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it.
\par
Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e. as $t$ increases). Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases. Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases. Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users. The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads users to have lower welfare than under no recommendation.

>>>>

\documentclass[sigconf]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage[font=small]{caption}
\usepackage{float}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

%
% end of the preamble, start of the body of the document source.
\begin{document}
% end of the preamble, start of the body of the document source.

@@

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
\end{abstract}

@@

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.

\maketitle

@@

Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets. For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption. Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.

Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations. While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead consumers into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.

\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we provide a theoretical model for how personalized recommender systems can shape the behavior of long-lived consumers and how the insights from our model can be used to improve recommender system design as well as allow us to better understand the consequences of these systems. In our model, consumers have uncertainty about their true utility of the goods in their choice set and consume only a small fraction of the large choice set they face.

First, we focus on understanding the degree to which recommender systems can lead consumers into \textit{filter bubbles}. Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}. The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated. This means that when consumers consume a good and learn its true utility this gives them information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty. This learning spill-over induces consumers to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns. This effect is further exemplified when consumers are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, consumers have a preference for goods with lower uncertainty to those with higher uncertainty.

We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns. We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences. We model existing recommendation systems as providing consumers with information on the common-value component and denote as this as partial recommendation. We consider how consumer behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for consumers if they knew the true utility of all the goods in their choice set.

Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst consumers.

\textbf{Related Work.}
Recommender Systems

@@

Viewing the goal of the RS as providing users with additional information they can utilize to update their beliefs, we first show that, without recommendation, users may get stuck in a ``filter bubble" whereby they consume increasingly similar content and that, in this case, welfare will be negatively correlated with diversity. We show how providing information to users via recommendation can lead users out of this filter bubble. However, if recommendation does not properly take into account user beliefs, recommendations can lead to increasing homogenization among users and lower welfare and consumption diversity. Finally we then show that when the recommender does not know the user's beliefs about her valuation, recommendation becomes decreasingly useful as the user learns more about her own preferences.
\par
\section{Our Model and Preliminaries}
\noindent \textbf{Expected Utility Theory} First, we review basic economic decision theory. [Duarte can do this]

\noindent \textbf{Consumers} We consider a set of consumers $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$. For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{N}$ the first time they consume it.

We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$. By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $u_{in}= v_{in} + \beta v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e. consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences.

@@

\noindent \textbf{Consumer Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item. We make this assumption as our model pertains to markets where the goods are experience goods and modeling consumers as being uncertain over their preferences in these environments  is common in the economics literature [add citations]. Even though consumers have access to online reviews for goods in many of the markets where recommender systems are deployed, it is prohibitively costly for consumers to acquire information on all of the goods in their choice set. Even if they did so, this information would be helpful for reducing uncertainty on the common-value component of their utility and not the idiosyncratic component. While we do not explicitly incorporate this into our model, we use it to motivate that consumers make consumption decisions under some degree of uncertainty.

Formally, consumer $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$. We impose the normality assumption for two reasons. The first is that consumers update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.

@@

\textbf{Consumer Learning}
When a consumer consumes a good $n$ they learn the realized utility for that good. In our model we incorporate the idea that learning the utility of good $n$ gives a consumer information about similar items. This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.

In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{N}$. This distance function has several nice properties.

To this effect, we consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
\par

@@

We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare. The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}. The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}. Note that a higher $H$ indicates more homogeneity. Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.

@@

Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$. In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility given beliefs at time $t$. Upon consuming that item and learning its value, the users update their beliefs about their valuation of other items and iterate this process $T$ times. In the second case, the recommender knows exactly $V_i$ and $V$ and can therefore recommend the best item for each consumer. In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$. The recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$. For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{

@@

\begin{table}[bt]
\centering
\begin{tabular}{rlll}
  \hline
Rec Policy & Welfare & Diversity & Homogeneity \\ 
  \hline
No Rec & 0.42 $\pm$ 0.002 & 0.19 $\pm$ 0.0003 & 0.001 $\pm$ 0 \\ 
Partial Rec & 1.25 $\pm$ 0.005 & 0.22 $\pm$ 0.0002 & 0.174 $\pm$ 0.004 \\ 
Omniscient Rec & 2.24 $\pm$ 0.006 & 0.24 $\pm$ 0.0001 & 0.035 $\pm$ 0.002 \\ 
   \hline
\end{tabular}
\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
\label{table:agg_results}
\end{table}
\begin{figure*}[!htbp]
    \centering
    \begin{minipage}[b]{0.5\textwidth}
        \centering
        \includegraphics[width=0.9\linewidth, height=1.5in]{local_moves_25}
        \caption{Extent of ``Local"  Consumption}
        \label{fig:local_consumption}
    \end{minipage}%
    ~ 
    \begin{minipage}[b]{0.5\textwidth}
        \centering
        \includegraphics[width=0.9\linewidth, height=1.5in]{welfare_homo_combo}
        \caption{Welfare and Homogeneity}
        \label{fig:beta_vary}
    \end{minipage}
    ~ \\
     \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=0.9\linewidth, height=1.5in]{rec_obedience_25}
     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
   \end{minipage}\hfill
   ~
   \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=0.9\linewidth, height=1.5in]{Diversity_vs_Welfare_No_Recommendation}
     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
   \end{minipage}
\end{figure*}

@@

\noindent \textbf{Results}. Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity. The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility. Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity. In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}. Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.  The partial recommendation case leads to higher consumption diversity and higher welfare, but also significantly higher homogeneity.

In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products. It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this. The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
\par
Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity. As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation. If recommendations only provides information on the common-value component (partially informed recommender) then welfare and consumption diversity increase when contrasted to the no recommendation case, but homogenization increases dramatically. Knowing the common-value component leads users to have more information on where to explore in the product space. However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users. If recommendation takes into account user beliefs on the idiosyncratic component of utility (omniscient recommender), then welfare and consumption diversity can further increase and homogeneity will decrease.
\par
Low consumption diversity is not necessarily bad by itself. Figure \ref{fig:diversity_welfare_no_rec} shows that, in the no recommendation case, welfare and consumption diversity are negatively correlated. This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products. However, if a user finds a high utility product right away, then staying in this neighborhood may yield higher utility due to the correlation of utilities. The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it due to having more information about these than more ``different'' ones, i.e. farther away.
\par
Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e. as $t$ increases). Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases. Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases. Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users. The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads to lower welfare than under no recommendation.

\section{The Value of the Ex-Ante View}

In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS. Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make \textit{ex-post} sub-optimal consumption decisions, but \textit{ex-ante} optimal given the information available to them at the moment of choice. Consider the following stylized example. The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs. As users obtain more information, their beliefs converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values. In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 2)$.
After uncertainty is unresolved, the user would consume $x_3$, but an expected utility maximizing consumer (without further information) would choose $x_1$.\footnote{We ignore an important component of decision-making under uncertainty, risk attitudes, by focusing on risk-neutral consumers. A \textit{risk-averse} user may choose to consume a good with lower expected \textit{ex-post} payoff but lower ex-ante variance compared to another good. While risk-aversion is important for decision-making under uncertainty, especially in the context of RS, it is not our main focus here.}
This observation, while seemingly trivial, has important implications. The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g. \cite[]{mas1995microeconomic}).} and allows the system designer to get some information about a user's beliefs. Why are user beliefs important objects for designers of RS to understand and utilize?
